### 2016-04-23 manuscript-body.tex
Fix typos; add Stefik text (need cite); add introduction

@@

\begin{enumerate}
  \tightlist
  \item When we use misconceptions to treat student ideas as right or wrong, we establish and perpetuate a deficit model of student understanding.
  \item The more classical misconceptions-basedresearch we publish as a community, the stronger that deficit narrative becomes in our discourse.
\end{enumerate}

This paper aims squarely at those problems by expanding a sense of what's possible in modeling cognition within computing education research. We argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have. First, we reanalyze prior data to show how we might recast misconceptions as the potential seeds of productive knowledge. We then argue through example that expertise in programming can involve the fluid and active construction of conceptual metaphors to deal with idiosyncrasies within and across programming languages. These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd. Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as knowledge and knowing in computing learning environments.

@@

The researchers are correct in the sense that \texttt{turkey} and
\texttt{ham} and \texttt{roastBeef} will \emph{not} be sorted into the
\texttt{meats} array. First, there is no code here that places
\texttt{turkey} and \texttt{ham} in the array; there is only code that
declares them as variables. Moreover, as written, an attempt to place
\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} into the array
would fail. Because of type restrictions in Java, only objects of class
Meat (or objects that inherit from class Meat) can go in the array.
\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} are, perhaps
confusingly, references to object instances of classes \texttt{Turkey}
and \texttt{Ham} and \texttt{RoastBeef}, so in the current snippet they
cannot enter an array of type \texttt{Meat} because (1) they don't yet
exist as objects and (2) even if they did exist, their types don't match
the array's type. The authors call this misconception \emph{semantics to
semantics}, which occurs ``when the student inappropriately
assume{[}s{]} details about the relationship and operation of code
samples, although such information was neither given nor implied''
\cite{kaczmarczyk_identifying_2010}.

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon. But, in their non-canonical thinking Student3 evidences
potentially ``\emph{productive}'' \cite{hammer_form_2002,hammer_tapping_2003} insights about design. Precisely \emph{because}
there is no code stating that \texttt{turkey} and \texttt{roastBeef} and
ham are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} could go
into an array of type \texttt{Meat}. To do so, a designer could define
\texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} as inheriting
from \texttt{Meat}.

@@

My second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's 1983 study of Pascal
programmers \cite{bonar_uncovering_1983}, data from which is also analyzed and discussed in \cite{pea_languageindependent_1986}. A student in
\cite{bonar_uncovering_1983} was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

@@

These two purposes, which for the sake of description I'll call
\emph{keeping control} (1a) and \emph{totaling up} (2a) are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

@@

An example drives home the connection between \cite{disessa_models_1986} and \cite{sherin_how_2001} as it applies to programming. The symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while base±change describes a kind of accumulation \cite{sherin_how_2001}.
Specifically, the terms in base±change ``contribute to a whole but play different roles. One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what I
would argue is the same symbolic structure: $\Box=︎ \Box + \Box$ (parts of a whole) and $\Box = \Box \pm \Delta$ (base + change) \cite{sherin_how_2001}.

@@

To return to misconceptions, what seems to drive research on students'
misconceptions is computing education largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
that's a worthwhile goals. But, as I've argued, we can
already identify cases where a narrow misconceptions focus is silent
about (at best) or dismissive of (at worst) students' useful intuitions. We identified
the further problems when a unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need. \emph{Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.} A knowledge-deficiency perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge as a problem. we begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

@@

Again, what's noteworthy is that rather than treating students'
non-canonical views as a burden for instruction, Bonar and Soloway
instead see them as an opportunity for programming language designers to
make languages better.

That insight---that instruction and design can meet novices where they
are---carries through to their final remarks about studying and
analyzing novice programming knowledge:

@@

 If we view the assignment statement as proposing
something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
field of knowledge and concomitant techniques from physics becomes
available to me to think with. But, if we focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what we know about physics in order to
reason about the code.

@@

Our second example concerns statements that use the same symbol template for fundamentally different kinds of computation. In our JavaScript example statements, all code came from the same language. But, another phenomenon comes
into play \emph{across} different languages when use the same symbology (written patterns) to stand for
conceptually different operations. The statements below are examples of what look
like semantically-equivalent operations, but in fact are not.

@@

Despite the syntactic similarity, The
layered creation of a plot invites a very different kind of conceptual
interpretation than, say, repeatedly accumulating numbers into a running
sum. One obvious reason for thinking about this code with a different
interpretive frame than incrementation is that numeric addition is commutative; composing a
plot is not necessarily commutative. So, despite the syntactic similarity, reassignments that
compose a plot using reassignment (as above) do not obey the same
rules as reassignments for a running total. But, it turns out the statements above don't obey the same rules of string concatenation either. Within this code
block, statements that look alike perform operations of a different
nature. While some expressions (lines 2 and 3) compose
visual layers onto a plot, others (lines 4 and 5) modify features of the plot, like the x and y scales.

@@

Resources can frames can interact in activity settings to produce
larger-scale patterns called ``epistemological coherences'' \cite{rosenberg_multiple_2006} where evidence from data suggests that a
network of discrete cognitive units can nonetheless give rise to stable
cognition.

>>>>

This paper aims to expand our sense of what's possible in modeling cognition within computing education research. We argue that research
approaches that privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have. We explore applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd. Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as \texttt{knowledge} and \texttt{knowing} in computing learning environments.

@@

\begin{enumerate}
  \tightlist
  \item When we use misconceptions to treat student ideas as right or wrong, we establish and perpetuate a deficit model of student understanding.
  \item The more classical misconceptions-based research we publish as a community, the stronger that deficit narrative becomes in our discourse.
\end{enumerate}

This paper aims squarely at those problems by expanding a sense of what's possible in modeling cognition within computing education research. We argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have. First, we reanalyze prior data to show how we might recast misconceptions as the potential seeds of productive knowledge. We then argue through example that expertise in programming can involve the fluid and active construction of conceptual metaphors to deal with idiosyncrasies within and across programming languages. These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition and symbolic forms can be a powerful framework for analysis in CSEd. Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as \texttt{knowledge} and \texttt{knowing} in computing learning environments.

@@

The researchers are correct in the sense that \texttt{turkey} and
\texttt{ham} and \texttt{roastBeef} will \emph{not} be sorted into the
\texttt{meats} array. First, there is no code here that places
\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} in the array; 
there is only code that
declares them as variables. Moreover, as written, an attempt to place
\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} into the array
would fail. Because of type restrictions in Java, only objects of class
\texttt{Meat} (or objects that inherit from class \texttt{Meat}) can go in the array.
\texttt{turkey} and \texttt{ham} and \texttt{roastBeef} are, perhaps
confusingly, references to object instances of classes \texttt{Turkey}
and \texttt{Ham} and \texttt{RoastBeef}, so in the current snippet they
cannot enter an array of type \texttt{Meat} because (1) they don't yet
exist as objects and (2) even if they did exist, their types don't match
the array's type. The authors call this misconception \emph{semantics to
semantics}, which occurs ``when the student inappropriately
assume{[}s{]} details about the relationship and operation of code
samples, although such information was neither given nor implied''
\cite{kaczmarczyk_identifying_2010}.

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon. But, in their non-canonical thinking Student3 evidences
potentially ``\emph{productive}'' \cite{hammer_form_2002,hammer_tapping_2003} insights about design. Precisely \emph{because}
there is no code stating that \texttt{turkey} and \texttt{roastBeef} and
\texttt{ham} are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} could go
into an array of type \texttt{Meat}. To do so, a designer could define
\texttt{Turkey} and \texttt{Ham} and \texttt{RoastBeef} as inheriting
from \texttt{Meat}.

@@

Our second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's 1983 study of Pascal
programmers \cite{bonar_uncovering_1983}, data from which is also analyzed and discussed in \cite{pea_languageindependent_1986}. A student in
\cite{bonar_uncovering_1983} was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

@@

These two purposes, which for the sake of description we'll call
\emph{keeping control} (1a) and \emph{totaling up} (2a) are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

@@

An example drives home the connection between \cite{disessa_models_1986} and \cite{sherin_how_2001} as it applies to programming. The symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while \emph{base+change} describes a kind of accumulation \cite{sherin_how_2001}.
Specifically, the terms in \emph{base+change} ``contribute to a whole but play different roles. One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what we
would argue is the same symbolic structure: $\Box=︎ \Box + \Box$ (parts of a whole) and $\Box = \Box \pm \Delta$ (base + change) \cite{sherin_how_2001}.

@@

To return to misconceptions, what seems to drive research on students'
misconceptions in computing education is largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
that's a worthwhile goal. But, as we've argued, we can
already identify cases where a narrow misconceptions focus is silent
about (at best) or dismissive of (at worst) students' useful intuitions. We identified
the further problems when a unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need. \emph{Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.} A knowledge-deficiency perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge as a problem. We begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

@@

Again, what's noteworthy is that rather than treating students'
non-canonical views as a burden for instruction, Bonar and Soloway
instead see them as an opportunity for programming language designers to
make languages better. Indeed, an empirical approach to language design has recently been articulated in Stefik and Siebert's work regarding language syntax's effect on novices. \cite{}

That insight---that instruction and design can meet novices where they
are---carries through to Bonar and Soloway's final remarks about studying and
analyzing novice programming knowledge:

@@

 If we view the assignment statement as proposing
something about the motion of an object---as, for example, when sense-making in science and engineering students can view equations as ``saying something'' about how a system behaves[blinded]---a
field of knowledge and concomitant techniques from physics become
available to me to think with. But, if we focus only on the
semantic-level equivalence of statements 1 through 5, there would be no
obvious reason for me to access what we know about physics in order to
reason about the code.

@@

Our second example concerns statements that use the same symbol template for fundamentally different kinds of computation. In our JavaScript example statements, all code came from the same language. But, another phenomenon comes
into play \emph{across} different languages when they use the same symbology (written patterns) to stand for
conceptually different operations. The statements below are examples of what look
like semantically-equivalent operations, but in fact are not.

@@

Despite the syntactic similarity, The
layered creation of a plot invites a very different kind of conceptual
interpretation than, say, repeatedly accumulating numbers into a running
sum. One obvious reason for thinking about this code with a different
interpretive frame than incrementation is that numeric addition is commutative; composing a
plot is not necessarily commutative. So, despite the syntactic similarity, statements that
compose a plot using reassignment (as above) do not obey the same
rules as reassignments for a running total. But, it turns out the statements above don't obey the same rules of string concatenation either. Within this code
block, statements that look alike perform operations of a different
nature. While some expressions (lines 2 and 3) compose
visual layers onto a plot, others (lines 4 and 5) modify features of the plot, like the x and y scales.

@@

Resources and frames can interact in activity settings to produce
larger-scale patterns called ``epistemological coherences'' \cite{rosenberg_multiple_2006} where evidence from data suggests that a
network of discrete cognitive units can nonetheless give rise to stable
cognition.
